---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# TDE-Dice

<!-- badges: start -->
<!-- badges: end -->

TDE-Dice brings the elegant probability mechanics of Germany's legendary RPG ["Das Schwarze Auge"](https://ulisses-spiele.de/game-system/das-schwarze-auge/) ([The Dark Eye]()https://ulisses-us.com/games/tde/) into (R)[https://www.r-project.org/]. Calculate exact success probabilities for attribute checks, combat manoeuvres, and skill checks—no more guessing whether your hero can scale that castle wall or parry the orc chieftain's axe!

Perfect for:
* Game Masters designing balanced encounters
* Statisticians exploring non-standard dice distributions
* Players optimizing character builds
* Educators teaching probability with RPG examples




## Core Mechanics Made Simple

The Dark Eye uses a unique d20 system with quality levels—where rolling under your attribute/skill value succeeds, and lower rolls yield better results. Our package handles the complex truncation and rectification rules automatically:

Key features: statistics for attribute, skill, and combat checks.<br/>

* R-style probability distributions for attributes, quality levels or skill points of skill checks, hit points of attack checks.
  * `d...()` for density functions.
  * `p...()` for distribution functions.
  * `q...()` for quantile functions.
  * `r...()` random generation functions.
* Functions to determine outcomes as success, fail, critical success, botch (with the name pattern `cSkill(...)`, `cAttr(...)` etc.).
* Basic helper functions for dice rolls to support custom calculations and comparisons (`dSumKdN()`, `convolveDice()`, `qualityLevel()` etc.).
* Plotting function (`plotAttributeChecks()`, `plotSkillChecks()`) based on [qqplot2](https://ggplot2.tidyverse.org/) supported by extensions of the probability functions to generate data frames.


## Comparative Checks

> *Under development*

What are the chances of **competitive** checks? Who is more likely to win? This can be used by game masters to tune the abilities and skill levels for NPCs.

What are the chances of **cumulative** checks? Tune opportunities to support your plot.

Chances of survival in combat. Give your health points and the weapon stats, which character is more likely to survive a fight? Or would your opponent escape the situation?


## Simulations

> *Under development*



## Example

This is a basic example which shows you how to solve common questions:


```{r example}
library(TDEDice)

# What's the chance a hero with Courage 14 succeeds 
# on a difficult check (target = 8) with 2 bonus dice?
cAttr(14)

# How good is this weapon? What are the chances of a 
# weapon hitting an opponent with a certain number of 
# hit points, excluding their defence?
dhitpoints(2:12, eav=12, w = c(Count=2, Dice=6, Mod=0))


# Visualize all possible outcomes for a sword-fighting check
#plotSkillDistribution(dSkillPoints(0:8, c(7, 10, 13), 8, "df"))
```




## Installation

You can install the development version of TDEDice from [GitHub](https://github.com/) with:

```r
# install.packages("pak")
pak::pak("SigurdJanson/TDE-Dice")
```

or with `devtools`:
```r
# install.packages("devtools")
devtools::install_github("SigurdJanson/TDE-Dice")
```

## ⚖️ License

MIT © 2026 Jan seifert. The Dark Eye (Das Schwarze Auge) is a registered trademark of Ulisses Spiele GmbH. This package is an unofficial fan tool and is not affiliated with or endorsed by Ulisses Spiele or Uhrwerk Verlag.


