---
title: "The Stats behind A Dark Eye"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The Stats behind A Dark Eye}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(TDEDice)
```


# Attributes

The [attribute check](https://tde.ulisses-regelwiki.de/checks.html):
* Roll a d20 against the effective attribute value (eav).
* Case 1 `d == 1`: a potential critical success is confirmed with another roll against the eav.
  * 1a. Confirmation roll `c <= eav` is a critical success.
  * 1b. Confirmation roll `c > eav` or `x == 20` is a common success.
* Case 2 `d == 20`: 

Hence, the effective attribute higher than 19 must still be treated like a 19 because the 20 is always considered a potential failure. An attribute of 20 still makes sense to compensate for penalties. But for the check it is limited.

## Critical Successes
$$
a = min(eav, 19) \newline
p_{1}(a) = \frac{1}{20} \cdot \frac{a}{20} =  \frac{a}{400}
$$

## Botches
$$
p_{20}(a) = \frac{1}{20} \cdot \frac{20-a}{20} = \frac{20-a}{400}
$$

## (Regular) Successes
$$
p_{s}(a) = \frac{a}{20} - p_{1}(a) = \frac{a}{20} -  \frac{a}{400} = \frac{20a}{400} -  \frac{a}{400} = \frac{19a}{400}
$$

## Failures
$$
p_{f}(a) = \frac{20-a}{20} - p_{20}(a) = \frac{20-a}{20} -  \frac{20-a}{400} = \frac{400-20a}{400} -  \frac{20-a}{400} = \frac{380 - 19a}{400}
$$



# Skill Checks

To identify the proper maths for skill checks we have to take into account that we treat a die roll below it's effective attribute value different from a roll above. Because in the end we are not interested in the dices faces. We are interested in the remaining skill points. When a result is above the eav, we lose skill points. But when it is below we do not gain any.

## Background

There are `r eavcombis <- 1540; eavcombis` combinations of 3 attribute values for a skill.

This is the result of choosing 3 values from 1 to 20 with replacement. Each eav may have the same value, hence, it must be chosen and the put back for the next eav.

Given:
$$
n = \text{number of distinct items to choose from} = 20  \newline
r = \text{number of items chosen} = 3
$$
This formula in R gives us the number of permutations:

```r
n <- 20; r <- 3L
factorial(n+r-1) / (factorial(r) * factorial(n-1))
#> [1] 1540
```
You can check it out here: https://www.graphcalc.com/combinations-replacement-calculator/.

Each of these `r eavcombis` combinations has a distinct probability distribution.

