---
title: "The Stats behind A Dark Eye"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The Stats behind A Dark Eye}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(TDEDice)
```


# Attributes

The [attribute check](https://tde.ulisses-regelwiki.de/checks.html):
* Roll a d20 against the effective attribute value (eav).
* Case 1 `d == 1`: a potential critical success is confirmed with another roll against the eav.
  * 1a. Confirmation roll `c <= eav` is a critical success.
  * 1b. Confirmation roll `c > eav` or `x == 20` is a common success.
* Case 2 `d == 20`: 

Hence, the effective attribute higher than 19 must still be treated like a 19 because the 20 is always considered a potential failure. An attribute of 20 still makes sense to compensate for penalties. But for the check it is limited.

## Critical Successes
$$
a = min(eav, 19) \newline
p_{1}(a) = \frac{1}{20} \cdot \frac{a}{20} =  \frac{a}{400}
$$

## Botches
$$
p_{20}(a) = \frac{1}{20} \cdot \frac{20-a}{20} = \frac{20-a}{400}
$$

## (Regular) Successes
$$
p_{s}(a) = \frac{a}{20} - p_{1}(a) = \frac{a}{20} -  \frac{a}{400} = \frac{20a}{400} -  \frac{a}{400} = \frac{19a}{400}
$$

## Failures
$$
p_{f}(a) = \frac{20-a}{20} - p_{20}(a) = \frac{20-a}{20} -  \frac{20-a}{400} = \frac{400-20a}{400} -  \frac{20-a}{400} = \frac{380 - 19a}{400}
$$


---
# Skill Checks

```{r SetupSkillBackground}

library(formattable)

library(jsonlite)
.attrs <- jsonlite::read_json(r"{D:\Programmierung\R\DSA\TDE-Dice\vignettes\attributes_en.json}")$Entries
dfattrs <- do.call(rbind, lapply(.attrs, as.data.frame))
rownames(dfattrs) <- NULL  # Remove row names

.skills <- jsonlite::read_json(r"{D:\Programmierung\R\DSA\TDE-Dice\vignettes\skills_en.json}")$Entries
dfskills <- do.call(rbind, lapply(.skills, as.data.frame))
rownames(dfskills) <- NULL  # Remove row names
```


To identify the proper maths for skill checks we have to take into account that we treat a die roll below it's effective attribute value different from a roll above. Because in the end we are not interested in the dices faces. We are interested in the remaining skill points. When a result is above the eav, we lose skill points. But when it is below we do not gain any.

## Background

There are `r eavcombis <- 1540; eavcombis` combinations of 3 attribute values for a skill.

This is the result of choosing 3 values from 1 to 20 with replacement. Each eav may have the same value, hence, it must be chosen and the put back for the next eav.

Given:
$$
n = \text{number of distinct items to choose from} = 20  \newline
r = \text{number of items chosen} = 3
$$
This formula in R gives us the number of permutations:

```r
n <- 20; r <- 3L
factorial(n+r-1) / (factorial(r) * factorial(n-1))
#> [1] 1540
```
You can check it out here: https://www.graphcalc.com/combinations-replacement-calculator/.

Each of these `r eavcombis` combinations has a distinct probability distribution.


## Attributes of Skill Checks

In total, there are `r length(.skills)` profane skills. Each skill requires 3 attributes for a check. Some attributes are needed more frequently for skill checks than others. The list shows how often an attribute is required in a skill check, with duplicate attributes in a single check counted twice.

```{r AttrUsePerSKill, echo=FALSE}
attrcount <- dfskills[, paste0("ab", 1:3)] |> 
  unlist() |>
  table() |> as.data.frame()
attrcount$Freq <- color_bar("lightgray")(attrcount$Freq)
attrcount$Label <- dfattrs$shortname
attrcount[, c("Var1", "Label", "Freq")] |>
  knitr::kable(col.names = c("Id", "Name", "Frequency"))
```

**Doubles**. In addition to the disproportionately frequent use of attributes, there is also an increasing number of instances where the same attribute is used twice in a skill check.

```{r AttrDoubleUsePerSkill, echo=FALSE}
attrcount <- dfskills[, paste0("ab", 1:3)]
# Find doubles
attrcount$doubles <- apply(attrcount, 1L, \(x) if (any(duplicated(x))) x[duplicated(x)] else NA_character_)

attrcount <- table(attrcount$doubles) |> as.data.frame()
for(a in dfattrs$attrID) {
  attrcount$Var1 <- sapply(attrcount$Var1, \(x) stringr::str_replace(x, a, dfattrs$shortname[dfattrs$attrID == a]))
}
attrcount$Freq <- color_bar("lightgray")(attrcount$Freq)
knitr::kable(attrcount, col.names = c("Combination", "Frequency"))
```


The repeated use of a single attribute in one skill check raises the question which attribute is needed for the largest set of skill checks? The INT attribute has the widest coverage, closely followed by SGC.

```{r MostUsedAttrAcrossSkill, echo=FALSE}
attrcount <- dfskills[, paste0("ab", 1:3)]
attrcount <- sapply(dfattrs$attrID, \(x) rowSums(attrcount[, paste0("ab", 1:3)] == x) > 0 |> sum())
attrcount <- colSums(attrcount) |> as.data.frame()

rownames(attrcount) <- NULL
attrcount$Name <- dfattrs$shortname
attrcount[[1]] <- color_bar("lightgray")(attrcount[[1]])
knitr::kable(attrcount[c(2, 1)], col.names = c("Name", "Frequency"))
```




**Triplets**. Finally, even triplets of all three attributes are noteworthily disbalanced. The triplet `SGC/SGC/INT` is the basis of 10 skill checks which makes almost `r round(10/59*100)`% of all skills.

```{r AttrTripleUsePerSKill-Calc, include=FALSE}
attrcount <- dfskills[, paste0("ab", 1:3)]
attrcount$triples <- apply(attrcount[, paste0("ab", 1:3)], 1L, \(x) paste0(x, collapse = "/"))

for(a in dfattrs$attrID) {
  attrcount$triples <- sapply(attrcount$triples, \(x) stringr::str_replace_all(x, a, dfattrs$shortname[dfattrs$attrID == a]))
}

attrcount <- table(attrcount$triples) |> as.data.frame()
```

```{r AttrTripleUsePerSKill-Table, echo=FALSE}
result <- attrcount[attrcount$Freq > 1L,]
result <- result[order(result$Freq, decreasing = TRUE),]
result$Freq <- color_bar("lightgray")(result$Freq)
rownames(result) <- NULL

knitr::kable(result, col.names = c("Combination", "Frequency"))
rm(result)
```

These `r length(attrcount[attrcount$Freq == 1L, "Var1"])` combinations are used just once `r paste(attrcount[attrcount$Freq == 1L, "Var1"])`.


