---
title: "The Stats behind The Dark Eye"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The Stats behind The Dark Eye}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(TDEDice)
library(ggplot2)
```


# Attributes

The [attribute check](https://tde.ulisses-regelwiki.de/checks.html):
* Roll a d20 against the effective attribute value (eav).
* Case 1 `d == 1`: a potential critical success is confirmed with another roll against the eav.
  * 1a. Confirmation roll `c <= eav` is a critical success.
  * 1b. Confirmation roll `c > eav` or `x == 20` is a common success.
* Case 2 `d == 20`: 

Hence, the effective attribute higher than 19 must still be treated like a 19 because the 20 is always considered a potential failure. An attribute of 20 still makes sense to compensate for penalties. But for the check it is limited.

## Critical Successes
$$
a = min(eav, 19) \newline
p_{1}(a) = \frac{1}{20} \cdot \frac{a}{20} =  \frac{a}{400}
$$

## Botches
$$
p_{20}(a) = \frac{1}{20} \cdot \frac{20-a}{20} = \frac{20-a}{400}
$$

## (Regular) Successes
$$
p_{s}(a) = \frac{a}{20} - p_{1}(a) = \frac{a}{20} -  \frac{a}{400} = \frac{20a}{400} -  \frac{a}{400} = \frac{19a}{400}
$$

## Failures
$$
p_{f}(a) = \frac{20-a}{20} - p_{20}(a) = \frac{20-a}{20} -  \frac{20-a}{400} = \frac{400-20a}{400} -  \frac{20-a}{400} = \frac{380 - 19a}{400}
$$


---
# Skill Checks

```{r SetupSkillBackground}

library(formattable)

library(jsonlite)
.attrs <- jsonlite::read_json(r"{D:\Programmierung\R\DSA\TDE-Dice\vignettes\attributes_en.json}")$Entries
dfattrs <- do.call(rbind, lapply(.attrs, as.data.frame))
rownames(dfattrs) <- NULL  # Remove row names

.skills <- jsonlite::read_json(r"{D:\Programmierung\R\DSA\TDE-Dice\vignettes\skills_en.json}")$Entries
dfskills <- do.call(rbind, lapply(.skills, as.data.frame))
rownames(dfskills) <- NULL  # Remove row names
```


To identify the proper maths for skill checks we have to take into account that we treat a die roll below it's effective attribute value different from a roll above. Because in the end we are not interested in the dices faces. We are interested in the remaining skill points. When a result is above the eav, we lose skill points. But when it is below we do not gain any.

## Background

There are `r eavcombis <- 1540; eavcombis` combinations of 3 attribute values for a skill.

This is the result of choosing 3 values from 1 to 20 with replacement. Each eav may have the same value, hence, it must be chosen and the put back for the next eav.

Given:
$$
n = \text{number of distinct items to choose from} = 20  \newline
r = \text{number of items chosen} = 3
$$
This formula in R gives us the number of permutations:

```r
n <- 20; r <- 3L
factorial(n+r-1) / (factorial(r) * factorial(n-1))
#> [1] 1540
```
You can check it out here: https://www.graphcalc.com/combinations-replacement-calculator/.

Each of these `r eavcombis` combinations has a distinct probability distribution.


## Attributes of Skill Checks

In total, there are `r length(.skills)` profane skills. Each skill requires 3 attributes for a check. Some attributes are needed more frequently for skill checks than others. The list shows how often an attribute is required in a skill check, with duplicate attributes in a single check counted twice.

```{r AttrUsePerSKill, echo=FALSE}
attrcount <- dfskills[, paste0("ab", 1:3)] |> 
  unlist() |>
  table() |> as.data.frame()
attrcount$Freq <- color_bar("lightgray")(attrcount$Freq)
attrcount$Label <- dfattrs$shortname
attrcount[, c("Var1", "Label", "Freq")] |>
  knitr::kable(col.names = c("Id", "Name", "Frequency"))
```

**Doubles**. In addition to the disproportionately frequent use of attributes, there is also an increasing number of instances where the same attribute is used twice in a skill check.

```{r AttrDoubleUsePerSkill, echo=FALSE}
attrcount <- dfskills[, paste0("ab", 1:3)]
# Find doubles
attrcount$doubles <- apply(attrcount, 1L, \(x) if (any(duplicated(x))) x[duplicated(x)] else NA_character_)

attrcount <- table(attrcount$doubles) |> as.data.frame()
for(a in dfattrs$attrID) {
  attrcount$Var1 <- sapply(attrcount$Var1, \(x) stringr::str_replace(x, a, dfattrs$shortname[dfattrs$attrID == a]))
}
attrcount$Freq <- color_bar("lightgray")(attrcount$Freq)
knitr::kable(attrcount, col.names = c("Combination", "Frequency"))
```


The repeated use of a single attribute in one skill check raises the question which attribute is needed for the largest set of skill checks? The INT attribute has the widest coverage, closely followed by SGC.

```{r MostUsedAttrAcrossSkill, echo=FALSE}
attrcount <- dfskills[, paste0("ab", 1:3)]
attrcount <- sapply(dfattrs$attrID, \(x) rowSums(attrcount[, paste0("ab", 1:3)] == x) > 0 |> sum())
attrcount <- colSums(attrcount) |> as.data.frame()

rownames(attrcount) <- NULL
attrcount$Name <- dfattrs$shortname
attrcount[[1]] <- color_bar("lightgray")(attrcount[[1]])
knitr::kable(attrcount[c(2, 1)], col.names = c("Name", "Frequency"))
```




**Triplets**. Finally, even triplets of all three attributes are noteworthily disbalanced. The triplet `SGC/SGC/INT` is the basis of 10 skill checks which makes almost `r round(10/59*100)`% of all skills.

```{r AttrTripleUsePerSKill-Calc, include=FALSE}
attrcount <- dfskills[, paste0("ab", 1:3)]
attrcount$triples <- apply(attrcount[, paste0("ab", 1:3)], 1L, \(x) paste0(x, collapse = "/"))

for(a in dfattrs$attrID) {
  attrcount$triples <- sapply(attrcount$triples, \(x) stringr::str_replace_all(x, a, dfattrs$shortname[dfattrs$attrID == a]))
}

attrcount <- table(attrcount$triples) |> as.data.frame()
```

```{r AttrTripleUsePerSKill-Table, echo=FALSE}
result <- attrcount[attrcount$Freq > 1L,]
result <- result[order(result$Freq, decreasing = TRUE),]
result$Freq <- color_bar("lightgray")(result$Freq)
rownames(result) <- NULL

knitr::kable(result, col.names = c("Combination", "Frequency"))
rm(result)
```

These `r length(attrcount[attrcount$Freq == 1L, "Var1"])` combinations are used just once `r paste(attrcount[attrcount$Freq == 1L, "Var1"])`.



## Skill Check Distributions

### Background

The starting point is the probability distribution for the sum of 3d20. The roughly bell-shaped form affects the logic. 

```{r SumOf3d20, echo=FALSE, fig.width=6}
df <- data.frame(
  Probability = d3D20(),
  DieRoll = 1:60L#,
  #Compensation = factor(c(rep("No", 3*8L), rep("Yes", 3*12L)))
)
ggplot(df, aes(x=DieRoll, y=Probability)) + #, fill = Compensation
  geom_col() +
  labs(title = "Probabilities for the Sum of 3d20" ) + #, subtitle = SubTitle)
  theme_minimal()
```



To approach the logic, let’s first assume that all three attributes have identical values. Let's say 8. This may rarely be the case in the game, but here it simplifies the logic enormously and makes it much easier to understand the principles behind DSA statistics. Why? When all 3 attributes are the same, we can simplify the calculation of the remaining skill points. We can compensate for a die roll that is higher than the corresponding attribute by using skill points. Now when all attributes have identical values, we can simply sum up all attributes for comparison. And we take the sum of the 3 die rolls except for one detail: every die roll lower than it's attribute uses the attribute value, otherwise we use the faces on the die. In code we use the sum of: `max(die, attribute)`. If we did not do that, A die roll below it's attribute would compensate die rolls above the attribute. But that does not happen in TDE.

The resulting distribution looks like this (bar chart: attributes 8/8/8 + skill 7). The line in the foreground shows the probabilities of the sum of 3d20 for comparison.


```{r echo=FALSE, fig.width=6}
df <- data.frame(
  p3d20 = d3D20(),
  pMaxSum = dSkillPurged(1:60, rep(8L, 3L), 7L, "vector") +
    crit3d20(rep(8L, 3L))/8000 + botch3d20(rep(8L, 3L))/8000,
  DieRoll = 1:60L#,
  #Compensation = factor(c(rep("No", 3*8L), rep("Yes", 3*12L)))
)
ggplot(df, aes(x=DieRoll, y=pMaxSum)) + #, fill = Compensation
  geom_col(fill = "lightgray") +
  geom_step(data = df, aes(x=DieRoll, y=p3d20), colour = "red") + 
  labs(title = "", caption = "Outcomes of a folded 3d20 roll.") + #, subtitle = SubTitle)
  theme_minimal()
```




```{r eval=FALSE, include=FALSE}
plotSkillDistribution(dSkillPoints(0:10, c(7, 10, 13), 8, "df"), 9)
```



## Skill Check Difficulty

Difficulty and quality are two different things, and yet, they are related. In many cases, we would assume that the need for a higher quality increases the challenge. But in the game, the game master would rarely add a penalty to a skill check **and** ask for a quality level above 1. That would mean increasing the challenge twice. Cases with a double challenge are magic spells that need to get through spirit (SPI) or toughness (TOU). The check is modified by the resistance of the target. Additionally, the quality level decides whether the spell is powerful enough to reach a desired goal, i.e. incapacitate the opponents, throw enough fire at them, or look into their minds long enough to gather important information.

Another good example for the double challenge are competitive checks. A penalty on the hero's check still requires them to out-roll the opponent with a higher quality level.

**Question**: what are the chances to reach a required quality level? Or in other words, when the game master requests a QL for success, how does it affect th character's performance? The following charts gives some answers. It shows the probability of each quality level depending on different configuration of attributes. To the right is the highest quality level. With a skill level of 18 that is QL6. But the curve of each single chart would not change much. The maximum QL on the right would roughly stay the same and we would lose the quality levels to the left when we reduce the character's skill level.

```{r SkillCheckByQL, echo=FALSE, fig.height=5, fig.width=10}
totalps <- NULL
Conditions <- list(Low = rep(5, 3L), Med = rep(10, 3L), High = rep(15, 3L))
# CREATE DATA FRAME
for (eav in Conditions) {
  print(names(eav))
  for (skill in seq(9L, 18L, 3L)) {
    for (mod in -1:1L) {
      ps <- dql(1:6, eav + mod, skill, format = "df")
      ps[ps[, "p"] == 0, "p"] <- NA_real_
      ps$EAV <- paste0(eav+mod, collapse = "/")
      ps$Skill <- skill
      ps$EAVRange <- unique(eav)
      ps$EAVMod <- mod
      
      totalps <- rbind(totalps, ps)
    }
  }
}
totalps$EAV <- ordered(totalps$EAV, levels=sapply(Conditions, \(x) paste0(x, collapse = "/")))
totalps$Skill <- as.ordered(totalps$Skill)
levels(totalps$QL) <- setNames(levels(totalps$QL), c("Failed", -5:-1, "Max"))

ggplot(totalps, aes(x=QL, y=p, group = Skill, colour = Skill)) +
  geom_line() + geom_point() +
  scale_color_discrete() +
  facet_grid(EAVRange ~ EAVMod, scales = "free_y") +
  geom_text(aes(label = EAV),
    x = -Inf, y = Inf,           # put text in top‑left corner
    hjust = -0.1, vjust = 1.1, size = 3L) +
  theme_minimal() |> 
  suppressMessages()
```

The TDE rule system clearly favours high quality levels over lower ones. While low attributes offer higher chances for low quality levels, higher attributes have **higher** chances on the **higher** quality levels.

Small variations have little effect on a character's success. A sequence such as 15/15/15 or 14/15/16 may be just as expensive to buy but it offer negligible advantages (not shown here). However, raising all attributes for a skill check by one has noteworthy effects on your chances.


**Question**: how do "incomplete" quality levels affect the probabilities?

When characters have skill points that can be divided by 3, they make the most of the highest skill level: that means they can use up to 2 skill points and they still get their maximum QL out of the check. That is the optimal case. How do chances change, when a character does not have 


```{r SkillCheckByQL_7_16, echo=FALSE, fig.height=5, fig.width=10}
totalps <- NULL
Conditions <- list(Med = rep(10, 3L), High = rep(15, 3L))
# CREATE DATA FRAME
for (eav in Conditions) {
  print(names(eav))
  for (skill in seq(9L, 18L, 1L)) {
    ps <- dql(1:6, eav, skill, format = "df")
    ps[ps[, "p"] == 0, "p"] <- NA_real_
    ps$EAV <- paste0(eav, collapse = "/")
    ps$Skill <- skill
    ps$EAVRange <- unique(eav)
    ps$Surplus <- TDEDice:::binSkillPoints(skill)[[1L]]

    totalps <- rbind(totalps, ps)
  }
}
totalps$EAV <- ordered(totalps$EAV, levels=sapply(Conditions, \(x) paste0(x, collapse = "/")))
totalps$Skill <- as.ordered(totalps$Skill)
#totalps$Surplus <- as.ordered(totalps$Surplus)
levels(totalps$QL) <- setNames(levels(totalps$QL), c("Failed", -5:-1, "Max"))

ggplot(totalps, aes(x = QL, y = p, group = Skill, colour = Skill, linewidth = Surplus)) +
  geom_line() + geom_point() +
  scale_linewidth(range = c(0.2, 1.2), breaks = sort(unique(totalps$Surplus))) + # balance widths
  scale_color_discrete() +
  facet_wrap(vars(EAVRange)) + #, scales = "free_y"
  geom_text(aes(label = EAV),
    x = -Inf, y = Inf,           # put text in top‑left corner
    hjust = -0.1, vjust = 1.1, size = 3L) +
  theme_minimal() |> 
  suppressMessages()
```


**Question**: how does an outlier in the attributes affect the chances?


```{r SkillCheckByQL_Outliers, echo=FALSE, fig.height=5, fig.width=10}
totalps <- NULL
Conditions <- list(L = c(5, 5, 5), LO = c(5, 5, 10), LOO = c(5, 5, 15),  
                   oM = c(10, 10, 5), M = c(10, 10, 10),  MO = c(10, 10, 15), 
                   ooH = c(15, 15, 5), oH = c(15, 15, 10), H = c(15, 15, 15))
# CREATE DATA FRAME
for (eav in Conditions) {
  for (skill in seq(9, 18, 3)) {
    ps <- dql(1:6, eav, skill, format = "df")
    ps[ps[, "p"] == 0, "p"] <- NA_real_
    ps$EAV <- paste0(eav, collapse = "/")
    ps$Skill <- skill
    #ps$EAVMax <- max(eav)
    #ps$Outlier <- 
    totalps <- rbind(totalps, ps)
  }
}
totalps$EAV <- ordered(totalps$EAV, levels=sapply(Conditions, \(x) paste0(x, collapse = "/")))
totalps$Skill <- as.ordered(totalps$Skill)
levels(totalps$QL) <- setNames(levels(totalps$QL), c("Failed", -5:-1, "Max"))

# PLOT
ggplot(totalps, aes(x=QL, y=p, group = Skill, colour = Skill)) +
  geom_line() + geom_point() +
  #scale_colour_manual(values = c("gray", "gray", "black")) +
  scale_color_discrete() +
  #ylim(0, 0.75) +
  facet_wrap(~ EAV, ncol=3, scales = "free_y") +
  theme_minimal()
```

**Question**: how does a penalty affect the probability of a skill check? From the previous charts we can already guess what will happen. Let us take a closer look at the following ones. Each chart shows a set of probabilities for a combination of attributes. Each data point shows the probability for a result of QL1 or higher.

```{r SkillCheckByMod, echo=FALSE, fig.height=5, fig.width=10}
totalps <- NULL
Conditions <- list(A = rep(8, 3L), AM = 7:9, AD = c(8, 8, 15), 
                   B = rep(11, 3L), BM = 10:12, BD = c(11, 11, 15), 
                   C = rep(15, 3L), CM = 14:16, CD = c(15, 15, 8))
skill <- 18

for (eav in Conditions) {
  for(mod in -5:0) {
    ps <- pql(1, eav+mod, skill, lower.tail = FALSE, format = "df")
    ps$EAV <- paste0(eav, collapse = "/")
    ps$Mod <- mod
    totalps <- rbind(totalps, ps)
  }
}
totalps$EAV <- ordered(totalps$EAV, levels=sapply(Conditions, \(x) paste0(x, collapse = "/")))
totalps$Skill <- skill
totalps$Skill <- as.ordered(totalps$Skill)

ggplot(totalps, aes(x=Mod, y=p, group = "black", colour = Skill)) +
  geom_line() + geom_point() +
  ylim(0, 1) + xlim(0, -5) +
  facet_wrap(~ EAV, ncol=3) +
  theme_minimal()
```


2. 

